Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-25 index.qmd
section:issues,measurement; process:write

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing. The CJ literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? and what is the solution? -->
However, the statistical literature cautions against using estimated scores for separate analyses and tests. A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power. This, in turn, increases the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010]. To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately. Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.

@@

<!-- Measurement model -->

<!-- a. measurement model is based on case V -->
<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
<!--   a.2 It does not consider that stimuli may be nested in individuals -->

<!-- Moreover, a notable concern in the second category is the measurement model's reliance on the assumptions of Case 5 from Thurstone's law of comparative judgment [-@Thurstone_1927]. As noted, it is common practice in the literature to use the BTL model as a means of generating CJ scores, serving as the measurement model. -->

<!-- Since the study by @Bramley_2008, it has been clear that the BTL model represents a statistical articulation of Case 5. -->

<!-- Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p. 269], its assumptions have become predominant in the literature due to its implementation through the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003 -->

<!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:  -->

<!-- (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects. -->

@@

<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind. This is studied by analysing the residuals that remain after the model is fitted to the data. Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large. Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability. indicates that one judge – Judge 5 – is significantly biased in some way. (Consider the problem of using adhoc methods) -->

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]. -->

<!-- (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. There is an amount of correlation that it is not accounted for.  -->
<!-- This is the case of clustering or measurement error, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.  -->

@@

<!-- (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.  -->

>>>>

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, researchers often use these BTL-generated scores or their transformations separately, conducting further analyses or hypothesis testing. The literature shows how these scores have been employed to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? and what is the solution? -->
However, the statistical literature cautions against using estimated scores for separate analyses and tests. A key consideration is that BTL-generated scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010]. To mitigate this risk, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately. Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.

@@

1. the BTL model is the statistical articulation of Case 5

  From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008

2. But this model hold the greater number of assumptions

  Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p. 269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions.

  @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three:

  (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects.

@@

<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind. This is studied by analyzing the residuals that remain after the model is fitted to the data. Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large. Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability. indicates that one judge – Judge 5 – is significantly biased in some way. (Consider the problem of using adhoc methods) -->

  (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for.

<!-- This is the case of clustering or measurement error, and overlooking issues such as clustering and measurement error can lead to biased and less precise parameter estimates [@McElreath_2020], ultimately diminishing the statistical power of models and increasing the likelihood of committing type I or type II errors when addressing research inquiries.  -->

@@

  (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals.

3. What problems does this bring?

<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
<!--   a.2 It does not consider that stimuli may be nested in individuals -->

  This leaves issues such as the presence of judge' biases hinted by @Bramley_2008 and @Kelly_et_al_2022, and evidenced by @Pollitt_et_al_2003, @Arlett_2003, and @Guthrie_2003

  [Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015]]{style="color:red;"}.
